---
title: "meuse_tutorial"
author: "nabil"
date: "January 20, 2015"
output: html_document
---


The following is taken from a vignette in the gstat package. gstat contains 
"geostatistical" functions, which can perform spatial (or, spatio-temporal) 
interpolations. This is particularly useful when you have data for certain 
points in a large area, and want some idea of how the values for that data 
being measured, might vary elsewhere.

In having tried to understand how work with spatial and spatio-temporal 
(hereafter denoted "S/T") data in R, I've realized there's a better way. First, 
for those looking to start, since gstat has functions that deal with spatial and 
S/T data, understanding how those structures work would probably be very useful 
when you start dealing with this package. 

At least for me, there were certain aspects of this tutorial which weren't 
obvious when I read through it, and in the end, took a while before I felt 
like I had a comfortable sense of what was happening. So, I thought I could 
elaborate on parts which weren't immediately obvious to me. Along the way, to 
help myself, I thought to code it how I've become used to. Hence, the plots 
are all re-done in ggplot2 (with the originals alongside for comparison), and 
the pipe operator %>% can really help clarify steps in involved operations. 

```{r}
library(sp)
library(tidyr)
library(ggplot2)
library(magrittr, warn.conflicts = FALSE)
```

Basically, the Meuse dataset contains measurements for concentrations of 
different elements, over an area in the Netherlands. Run ?meuse for more info.

```{r}
data(meuse)
class(meuse)
str(meuse)
```

If you'll notice, there are two columns "x" and "y" in the dataset. These are 
the coordinates of the location which that row of the dataframe corresponds to 
(if the numbers seem large, don't worry, it's just Rijksdriehoek (RDH) 
coordinates, used in the Netherlands.)

Now, while it's easy to include spatial information, like coordinates, just 
in columns of a dataframe, if someone else looks at the data, it might not be 
obvious what all the variables in the columns represent. So, one way to deal 
with this is to more concretely identify the observations with the corresponding 
location. Here, the result is going to be a Spatial Points Data Frame. This 
assignment can be done like normally, except that the formula notation is used:

```{r}
coordinates(meuse) <- ~ x + y
class(meuse)

```

Once that assignment is performed, then the class of the object changes 
accordingly.

Behind the scenes, this new Spatial*DataFrame is actually an S4 object (i.e., 
that's the system of OO-programming used for it in R) :

```{r}
str(meuse)
```

While accessing elements of such an objects "slots" is discouraged, the sp 
package comes with useful helper functions that can be used, such as bbox and 
coordinates. The summary function also has slightly different output: 

```{r}
meuse %>% coordinates %>% head
meuse %>% bbox
summary(meuse)
```

Typically, to just access the data, I coerce the object to a dataframe with 
as.data.frame, which can be a lot quicker and cleaner than alternatives:

```{r, eval=FALSE}
meuse_df <- cbind( attr(meuse, "data"), meuse@coords) # just coerce to df
```

The sp package comes with special built-in graphing functions, such as "bubble": 

```{r}
# bubble chart
bubble(meuse, "zinc", col = c("#00ff0088", "#00ff0088"), 
       main="zinc concentrations (ppm)")
```

The equivalent can also be made using ggplot:

```{r}
# I think the blue stands out better against the white background
meuse %>% as.data.frame %>% 
  ggplot(aes(x, y)) + geom_point(aes(size=zinc), color="blue", alpha=3/4) + 
  ggtitle("Zinc Concentration (ppm)") + coord_equal() + theme_bw()
```


```{r}
## Project the data from Rijksdriehoek (RDH) (Netherlands topographical) map 
## coordinates to Google Map coordinates; RDH coordinates have an EPSG code of 
## 28992 and Google map coordinates have an EPSG code of 3857

# But from the documentation of proj4string: Note that only “+proj=longlat” is 
# accepted for geographical coordinates, which must be ordered (eastings, 
# northings). So use sp

# plan: convert rdh to longlat, then assign longlat, then transform to rdh
# TODO: incorporate this into a post on using ggmap with spatial data.
library(rgdal)

ESPG <- make_EPSG()
ESPG[which(ESPG$code == 28992), ]
rdh_proj <- ESPG[which(ESPG$code == 28992), "prj4"]

#proj4string(meuse) = "+proj=longlat +datum=WGS84"
```

Along with the meuse dataset is one called meuse.grid. Later on in the 
interpolation, it's used as locations to predict concentrations for.  At first, 
it's just a regular dataframe like meuse was:

```{r}
data(meuse.grid)
summary(meuse.grid)

meuse.grid %>% str
meuse.grid %>% class
```

To better see the gridded nature of the data, we can just plot it:

```{r}
# this is clearly gridded over the region of interest
meuse.grid %>% as.data.frame %>% 
  ggplot(aes(x, y)) + geom_point(size=1) + coord_equal()
 
# to compare, recall the bubble plot above; those points were what there were 
# values for. this is much more sparse
meuse %>% as.data.frame %>% 
  ggplot(aes(x, y)) + geom_point(size=1) + coord_equal()
```

These two plots pretty much summarize our interpolation problem: given values 
at the locations in the latter plot, we want to interpolate over all values in 
the former plot. 

And just as before, we specify that the x and y columns are actually coordinates 
for the observations. Here, though, we can also manually specify that meuse.grid 
actually cotains a grid of points. Although this might not appear to change 
anything if you only inspect the class, the attributes of the object do change 
(to see that, just check the attributes before and after identifying it as 
gridded).

```{r}
coordinates(meuse.grid) = ~x+y
gridded(meuse.grid) = TRUE
meuse.grid %>% class
```

More plotting

```{r}
image(meuse.grid["dist"])
title("distance to river (red=0)")

# ggplot version
meuse.grid %>% as.data.frame %>%
  ggplot(aes(x, y)) + geom_tile(aes(fill=dist)) + 
  scale_fill_gradient(low = "red", high="yellow") + coord_equal() + theme_bw() + 
  ggtitle("Distance to River")
```

Alternatively, instead of using tiles, one could go Seurat-style and call 
geom_point() with small size (but, just note that scale_color_gradient goes with 
points, and scale_fill_gradient with tiles). 

## Kriging interpolation of remaining points

To recap up to this point: we have values at some points, and want to 
interpolate over an entire grid. In this case, we can use gstat's kriging 
functions. In particular, we'll just start off with the simple "krige" for now.

```{r}

library(gstat)

zinc.idw <- krige(zinc ~ 1, meuse, meuse.grid)
zinc.idw %>% class
zinc.idw %>% as.data.frame %>% head
```

Here, there are a couple things to note. First, the function takes a "formula" 
argument. Since we want to interpolate for values of zinc, we would use 
"ordinary", or "simple", kriging, in which case we use the notation 
"[variable] ~ 1". The second argument is the where the values of that 
variable being interpolated, come from. The third is the region of interest, 
such as a grid of spatial locations we want estimated predictions for.

The result of the kriging is a data frame with coordinates (x and y), 
predicted values of the variable ("var1.pred"), and variance of the 
predictions ("var1.var"). (Aside: I'm not really sure why in this example, there are NA's 
for the prediction variance; I think it's because a variogram wasn't supplied 
to form the predictions from. However, I'm almost not clear how there can be 
predictions without that variogram, but I haven't studied much of the theory 
behind this yet).

These results, again, can be graphed with the sp package's functions, or 
otherwise with ggplot2:

```{r}
spplot(zinc.idw["var1.pred"], main="zinc inverse distance weighted interpolations")

#same spplot with ggplot 
library(scales)

zinc.idw %>% as.data.frame %>% 
  ggplot(aes(x=x, y=y, fill=var1.pred)) + geom_tile() + theme_bw() + 
  coord_equal() + scale_fill_gradient(low = "red", high="yellow") + 
  ggtitle("zinc inverse distance weighted interpolations") + 
  scale_x_continuous(labels=comma) + scale_y_continuous(labels=comma) 
```

One advantage of ggplot2 in this case is the amount of control over the color 
scheme (as well as other aspects of the plot). In the example above, I stuck 
to the red-to-yellow scale used earlier. Although, note that here, red doesn't 
represent "0" anymore.

```{r}
# graphical check of hypothesis from above graphs
plot(log(zinc) ~ sqrt(dist), data=meuse, pch=16, cex=.5)
abline(lm(log(zinc) ~ sqrt(dist), meuse))

# or with ggplot:
# meuse %>% as.data.frame %>% 
#   ggplot(aes(sqrt(dist), log(zinc))) + geom_point() + 
#   geom_smooth(method="lm", se=FALSE)
```



## Variogram plotting 

As alluded to earlier, it's often helpful when performing kriging to also have 
a variogram (or, semi-variogram) model fit to the data. For an excellent 
introduction to variograms, see 
[Allison Lassiter's website](http://allisonlassiter.com/2014/05/13/kriging-overview-part-1-variograms/).

Basically, while we did interpolate using a relationship for zinc, we might 
want to explore how log(zinc) varies over space. For this, we can plot a 
variogram. First, the code, then more explanation about what the code does.

```{r}
# inspect variation of log(zinc) by distance (i.e., from the river)
lzn.vgm <- variogram(log(zinc)~1, meuse) # calculates sample variogram values
lzn.fit <- fit.variogram(lzn.vgm, model=vgm(1, "Sph", 900, 1)) # fit model
plot(lzn.vgm, lzn.fit) # plot the sample values, along with the fit model
```

In the first line, we merely calculate a sample variogram. This involves 
several things, as can be seen by inspecting the actual object: 

```{r}
lzn.vgm
lzn.vgm %>% class
lzn.fit %>% class
```

The first column, np, says how many point pairs were within distance "dist" (if 
those numbers look like a lot, recall that although meuse has only 155 rows, 
there are 155 * 154 / 2 = 11,935 point pairs; see the plot above with 
the points graphed). If we plot this object itself, we just get the sample 
variogram, without any fit to it (try it!). 

To perform a fit, we call the fit.variogram function, and pass it two 
parameters: a variogram object, and a model we want to fit the data to. With the 
model specified, the function would find the optimal (in some sense) parameters 
for that model to fit  the data. 

In this tutorial, a spherical model is used. The book Applied Spatial Data Analysis 
with R (ASDAR) has the complete list of variogram models one can use. And while 
the functional forms of those models aren't included, a more 
graphical/qualitative display of characteristics for different variogram models, 
is available by calling the function: show.vgms(). 

Now if you plot the variogram and the fit, you (surprise!) get both together. 

But we might not like that model. So we could try to see how log(zinc) varies 
with the square root of distance. This time, we'll try an exponential model. 
Otherwise, everything is pretty much the same as before:

```{r}
# inspect variation of log(zinc) by square root of distance 
lznr.vgm <- variogram(log(zinc) ~ sqrt(dist), meuse)
lznr.fit <- fit.variogram(lznr.vgm, model=vgm(1, "Exp", 300, 1))
lznr.fit %>% class
plot(lznr.vgm, lznr.fit)
```

## Section 6: Kriging

 question: how do kriging results vary if no model specified?
 question: how does kriging happen when no projection specified? This seems 
           to be opposed to the meuse tutorial.
 note: here, interpolation done on gridded SPDF, but this time, result is 
       another SPDF (unlike when not specifying vgm model). Also, there are 
       values for var1.var in the output (which seems to be variance of 
       the prediction). 

```{r}
lzn.kriged <- krige(log(zinc) ~ 1, meuse, meuse.grid, model=lzn.fit)
```



```{r}
# sp plotting
spplot(lzn.kriged["var1.pred"])
```



```{r}
# kriging results in ggplot
lzn.kriged %>% as.data.frame %>% 
  ggplot(aes(x=x, y=y)) + geom_tile(aes(fill=var1.pred)) + 
  coord_equal() + scale_fill_gradient(low = "red", high="yellow") + 
  scale_x_continuous(labels=comma) + scale_y_continuous(labels=comma) + 
  theme_bw() 

```



## Section 7: Conditional Simulations

```{r}
lzn.condsim <- krige(log(zinc)~1, meuse, meuse.grid, model=lzn.fit, 
                     nmax=30, nsim=4)
# sp plotting
spplot(lzn.condsim, main="three conditional simulations")
```


```{r}
# with ggplot2. (no need to call components with "@" or "attr(., "data"), e.g.) 
#lzn_cond_df <- cbind(attr(lzn.condsim, "data"), attr(lzn.condsim, "coords"))
lzn.condsim %>% as.data.frame %>% 
  gather(sim, value, sim1:sim4) %>% 
  ggplot(aes(x=x, y=y)) + geom_tile(aes(fill=value)) + 
  facet_grid(.~sim) + coord_fixed(ratio = 1) + 
  scale_x_continuous(labels=comma) + scale_y_continuous(labels=comma) + 
  scale_fill_gradient(low = "red", high="yellow") + 
  ggtitle("Three conditional simulations") + theme_bw()
```






